Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-09-05 manuscript/draft.Rmd
cite Ward in discussion. fixes #190
fix blonder citation per Shawn's advice

We have only evaluated annual forecasts up to a decade into the future, but
forecasts are often made with a lead time of 50 years or more. These long-term
forecasts are difficult to evaluate given the small number of century-scale
datasets, but are important for understanding changes in biodiversity at some of
the lead times relevant for conservation and management. Two studies have
assessed models of species richness at longer lead times [@algar2009;
@distler2015], but the results were not compared to baseline or time-series
models (in part due to data limitations) making them difficult to compare to our
results directly. Studies on shorter time scales, such as ours, provide one way
to evaluate our forecasting methods without having to wait several decades to
observe the effects of environmental change on biodiversity [@petchey2015;
@tredennick2016; @dietze2016], but cannot fully replace longer-term evaluations
[@tredennick2016]. In general, drivers of species richness can differ at
different temporal scales [@rosenzweig1995; @white2004; @white2007], so
different methods may perform better for different lead times. In particular, we
might expect environmental and ecological information to become more important
at longer time scales, and thus for the performance of simple baseline forecasts
to degrade faster than forecasts from SDMs and other similar models
[@blonder2017].  We did observe a small trend in this direction: deviance for
the auto-ARIMA models and for the average baseline grew faster than for two of
the environmental models (the JSDM and the macroecological model), although this
growth was not statistically significant for the average baseline.

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines that assume the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness.  The
site-level stability of the BBS data also explains why SDMs and macroecological
models perform relatively well at predicting future richness, despite failing to
capture changes in richness over time. However, this stability also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. These results suggest that single-site models should be actively
considered for forecasts of richness and other stable aspects of biodiversity.
Our results also suggest that future efforts to understand and forecast
biodiversity should incorporate species composition, since lower-level processes
are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
useful information [@harris2015].

>>>>

We have only evaluated annual forecasts up to a decade into the future, but
forecasts are often made with a lead time of 50 years or more. These long-term
forecasts are difficult to evaluate given the small number of century-scale
datasets, but are important for understanding changes in biodiversity at some of
the lead times relevant for conservation and management. Two studies have
assessed models of species richness at longer lead times [@algar2009;
@distler2015], but the results were not compared to baseline or time-series
models (in part due to data limitations) making them difficult to compare to our
results directly. Studies on shorter time scales, such as ours, provide one way
to evaluate our forecasting methods without having to wait several decades to
observe the effects of environmental change on biodiversity [@petchey2015;
@tredennick2016; @dietze2016], but cannot fully replace longer-term evaluations
[@tredennick2016]. In general, drivers of species richness can differ at
different temporal scales [@rosenzweig1995; @white2004; @white2007;
@blonder2017], so different methods may perform better for different lead times.
In particular, we might expect environmental and ecological information to
become more important at longer time scales, and thus for the performance of
simple baseline forecasts to degrade faster than forecasts from SDMs and other
similar models.  We did observe a small trend in this direction: deviance for
the auto-ARIMA models and for the average baseline grew faster than for two of
the environmental models (the JSDM and the macroecological model), although this
growth was not statistically significant for the average baseline.

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines that assume the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness. Ward et al.
(2014) found similar patterns in time series of fisheries stocks, where
relatively stable time series were best predicted by simple models and more
complex models were only beneficial with dynamic time series. The site-level
stability of the BBS data also explains why SDMs and macroecological models
perform relatively well at predicting future richness, despite failing to
capture changes in richness over time. However, this stability also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. These results suggest that single-site models should be actively
considered for forecasts of richness and other stable aspects of biodiversity.
Our results also suggest that future efforts to understand and forecast
biodiversity should incorporate species composition, since lower-level processes
are expected to be more dynamic [@ernest2001; @dornelas2014] and contain more
useful information [@harris2015].
